Consistent Polyhedral Surrogates via Embedding

### 2019-09-10 notes/top-k.tex
Updating top k to have embedded loss instead of property conjecture

\section{Top-$k$ surrogate}
Throughout this section, consider the surrogate and discrete loss $L^{top-k}(u)_y~=~\left(\frac{1}{k} \sum_{i=1}^k (u + \ones - e_y)_{[i]} - u_y \right)_+$.

First, we want to write $L^{top-k}$ in a linear form with the $\max(\cdot, 0)$ operation.
Then, using this linear form of the expected loss, we are able to evaluate when certain report values are optimal for the expected loss as a function of $p$.

\subsection{Notation and Assumptions}
\subsubsection{Notation}
Let $\bar{u} = \frac 1 k \sum_{i = 1}^k u_i$ and similarly, $\bar{u}_{-y} = \frac{1}{k-1} \sum_{i=1, i \neq y}^k u_i$ be the averages of the first $k$ elements of $u$ and the average of the first $k$ elements of $u$ besides $y$, respectively.
Moreover, we say $b = \argmax_i \{u_i : u_i > 1\}$.

We use $\odot$ to denote the Hadamard (element-wise) vector product.
Also, $\min(\vec u, \vec v)$ is the element-wise minimum of vectors $\vec u, \vec v \in \reals^d$.

\subsubsection{Assumptions}
Throughout, we make two assumptions for ease of exposition, but both follow without loss of generality for an optimal report.
First, assume that $u$ is ordered, where $u_1 \geq u_2 \geq \ldots \geq u_n$.
This is without loss of generality, as we can simply reorder $p$ if this is not the case.

@@

Observe that, for a minimizer of $L^{top-k}$, we can assume there is no $u_i$ such that $u_i > \bar u_{-i} +1$, since reassigning such a $u'_i := \bar u_{-i} +1$ results in $L^{top-k}(u')_i = L^{top-k}(u)_i = 0$, but reduces $L^{top-k}(\cdot)_j$ for all $j \neq i$ by decreasing $\ubar$.

To see this, consider $u_y = \bar u_{-y} + 1 + \epsilon$ and $u'_y = \bar u_{-y} + 1$, with $u_i = u'_i \; \forall i \neq y$. 
\begin{align*}
L^{top-k}(u)_y &= \left(1 - (\bar u_{-y} + \epsilon + 1) + \frac 1 k \left(\sum_{i=1, i \neq y}^k u_i + \bar u_{-y} + \epsilon\right) \right)_+\\ 
&= \left(1 - (\bar u_{-y} + \epsilon + 1) + \frac 1 k \left((k-1) \bar u_{-y} + \bar u_{-y} + \epsilon\right) \right)_+\\
&= \left(-\bar u_{-y} - \epsilon + \bar u_{-y} + \frac \epsilon k \right)_+ \\
&= (-\frac{k-1}{k}\epsilon)_+\\
&= 0
\end{align*}
\begin{align*}
L^{top-k}(u')_y &= \left(1 - (\bar u_{-y} + 1) + \frac 1 k \left(\sum_{i=1, i \neq y}^k u_i + \bar u_{-y} \right) \right)_+\\ 
&= \left(1 - (\bar u_{-y} + 1) + \frac 1 k \left((k-1) \bar u_{-y} + \bar u_{-y} \right) \right)_+\\
&= \left(-\bar u_{-y} + \bar u_{-y}\right)_+ \\
&= 0
\end{align*}

Now, for any outcome $i \neq y$, we know that $\bar u_{-i} > \bar u_{-i}$ because of $u_y$.
Since $u_y$ is only added in $L^{top-k}(\cdot)_i$ as part of the $\bar u_{-i}$ term, we can see that $L^{top-k}(u)_i \geq L^{top-k}(u')_i$ for all $i \neq y$.

@@

First, observe that we can re-write
\begin{align*} 
L^{top-k}(u)_y&=\left(\frac{1}{k} \sum_{i=1}^k (u + \ones - e_y)_{[i]} - u_y \right)_+\\ 
&=\left(1 - u_y + \frac{1}{k} \sum_{i=1}^k (u - e_y)_{[i]} \right)_+\\
&=\left(1 - u_y + \frac{1}{k} \sum_{i=1}^k (u - \min(u_y, 1)\Ind{y \leq k})_{i} \right)_+
\end{align*}
\jessie{Do we need the indicator given the assumptions...?}

Consider that the loss value is $1 + \bar u$ if $y > k$, since $u_y = 0$ in this case, so the loss is nonnegative even before applying the $(\cdot)_+$ operator.
If $y \leq k$ and $u_y \in [0,1]$, then the loss is nonnegative before applying $(\cdot)_+$ as $1-u_y \geq 0$, as is $\bar u$.
Now if $u_y \in (1, \bar u_{-y} + 1]$, then the term inside $(\cdot)_+$ of $L^{top-k}(u)_y$ is equal to $\frac{1}{k-1} + 1 +  \ubar_{-y} - u_y $
Since $u_y \leq 1 +  \ubar_{-y}$, we can see this term is nonnegative.
Therefore, we can drop the $(\cdot)_+$ operator on this report space.

We can now write the expected loss as follows:
\begin{align*}
\inprod{p}{L^{top-k}(u)} &= \sum_{y=1}^n p_y L^{top-k}(u)_y \\
&= \sum_{y=1}^b p_y L^{top-k}(u)_y + \sum_{y=b+1}^k p_y L^{top-k}(u)_y + \sum_{y=k+1}^n p_y L^{top-k}(u)_y\\
&= \sum_{y=1}^b p_y (1 - \frac 1 k - u_y + \ubar) + \sum_{y=b+1}^k p_y (1 - \frac{k+1} k u_y + \ubar) + \sum_{y=k+1}^n p_y(1-u_y + \ubar)\\
&= 1 + \ubar - \inprod{p}{u} - \inprod{p \odot \Ind{y \leq i}}{\frac 1 k \ones} - \inprod{p \odot \Ind{i+1 \leq y \leq k}}{ \frac u k}\\
&= 1 + \ubar - \inprod{p}{u + \frac 1 k \min(u, \ones)}
\end{align*}
When considering the Bayes Risk, we use this final form of the expected loss.

\subsection{Bayes Risk}
First, consider the case where $u_y = 0$ and $u'_y = 1$, with $u_j = u'_j \;\forall i \neq j$.
As the expected loss is linear on $u_y$, and these are the boundary cases, we know that $u$ is optimal on one of these two reports in this case.

So we ask when is $\inprod{p}{L^{top-k}(u)} \leq \inprod{p}{L^{top-k}(u')}$?
\begin{align*}
1 + \ubar - \inprod{p}{u + \frac 1 k \min(u, \ones)} &\leq 1 + \ubar' - \inprod{p}{u' + \frac 1 k \min(u', \ones)}\\
- \inprod{p}{u + \frac 1 k \min(u, \ones)} &\leq \frac {1}{k} - \inprod{p}{u' + \frac 1 k \min(u', \ones)}\\
0 &\leq \frac{1}{k} - p_y(\frac{k+1}{k})\\
p_y &\leq \frac{1}{k+1}
\end{align*}

Conversely, if $p_y > \frac{1}{k+1}$, it is to better to report $u_y = 1$ than $u_y = 0$.
However, since we are not restricted to the unit hypercube, we might ask when it is optimal to bring $u_y = 1$ up to $u'_y = 1 + \epsilon$ given $\epsilon > 0$.
Here, we consider when $\inprod{p}{L^{top-k}(u)} \leq \inprod{p}{L^{top-k}(u')}$.
\begin{align*}
1 + \ubar - \inprod{p}{u + \frac 1 k \min(u, \ones)} &\leq 1 + \ubar' - \inprod{p}{u' + \frac 1 k \min(u', \ones)}\\
\frac{1}{k} - p_y (\frac{k+1}{k}) &\leq \frac{1+\epsilon}{k} - p_y (\frac{k+1}{k}(1 + \epsilon))\\
p_y &\leq \frac{1}{k}
\end{align*}
In this case, $u_y=1$ is preferable to $u+y >1$.
If $p_y > \frac 1 k$, then we want to push $u_y$ as high as we can, but recall that we restrict to $u_y \leq \ubar_{-y} + 1$.
Because the upper bound on $u_y$ depends on $\ubar$, we can't determine $u$ iteratively.
However, we can construct $u$ as given below.
\begin{conjecture}
	Fix $p \in \simplex$.
	Let $g_t = |\{p_i : p_i \geq \frac 1 k\}|$ and $g_b = | \{p_i \in (\frac{1}{k+1}, \frac{1}{k})\}|$.
	The property \[\gamma(p) = 
	\left[
	\begin{cases}
	0 & p_i \leq \frac {1}{k+1}\\
	1 & p_i \in (\frac{1}{k+1}, \frac{1}{k})\\
	\frac{g_b + k-1}{k-g_t} & o/w
	\end{cases}
	\right]_{i=1}^n
	\] refines the property $\prop{L^{top-k}}$.
\end{conjecture}
\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

>>>>

\section{Top-$k$ surrogate}
Throughout this section, consider the surrogate and discrete loss \begin{equation}\label{eq:top-k-surrogate}
L^{top-k}(u)_y~=~\left(1 - u_y + \frac{1}{k} \sum_{i=1}^k (u - e_y)_{[i]} \right)_+
\end{equation}

First, we want to write $L^{top-k}$ in a linear form with the $(\cdot)_+$ operator.
Then, using this linear form of the expected loss, we are able to write a finite set of optimal reports over $\simplex$.

\subsection{Notation and Assumptions}
\subsubsection{Notation}
Let $\bar{u} = \frac 1 k \sum_{i = 1}^k u_i$ and similarly, $\bar{u}_{-y} = \frac{1}{k-1} \sum_{i=1, i \neq y}^k u_i$ be the averages of the first $k$ elements of $u$ and the average of the first $k$ elements of $u$ besides $y$, respectively.
Moreover, we say $b = \argmax_i \{u_i : u_i > 1\}$, assuming $u$ is ordered.
See below for justification.

We use $\odot$ to denote the Hadamard (element-wise) vector product.
Also, $\min(\vec w, \vec v)$ is the element-wise minimum of vectors $\vec w, \vec v \in \reals^d$.

\subsubsection{Assumptions}
Throughout, we make two assumptions for ease of exposition, but both follow without loss of generality for an optimal report.  %, as well as a third assumption when thinking about minimizing the expected loss.
First, assume that $u$ is ordered, where $u_1 \geq u_2 \geq \ldots \geq u_n$.
This is without loss of generality, as we can simply reorder $p$ if this is not the case.

@@

\begin{lemma}\label{lem:top-threshold}
	There is a minimizer $u$ of $L^{top-k}$ such that for all $u_i$, we have $u_i \leq \bar u_{-i} +1$. 
\end{lemma}
\begin{proof}
	Reassigning such a $u'_i := \bar u_{-i} +1$ results in $L^{top-k}(u')_i = L^{top-k}(u)_i = 0$, but reduces $L^{top-k}(\cdot)_j$ for all $j \neq i$ by decreasing $\ubar$.

To see this, consider $u_y = \bar u_{-y} + 1 + \epsilon$ and $u'_y = \bar u_{-y} + 1$, with $u_i = u'_i \; \forall i \neq y$. 
\begin{align*}
L^{top-k}(u)_y &= \left(1 - (\bar u_{-y} + \epsilon + 1) + \frac 1 k \left(\sum_{i=1, i \neq y}^k u_i + \bar u_{-y} + \epsilon\right) \right)_+\\ 
&= \left(1 - (\bar u_{-y} + \epsilon + 1) + \frac 1 k \left((k-1) \bar u_{-y} + \bar u_{-y} + \epsilon\right) \right)_+\\
&= \left(-\bar u_{-y} - \epsilon + \bar u_{-y} + \frac \epsilon k \right)_+ \\
&= (-\frac{k-1}{k}\epsilon)_+\\
&= 0
\end{align*}
Compare the above to the following:
\begin{align*}
L^{top-k}(u')_y &= \left(1 - (\bar u_{-y} + 1) + \frac 1 k \left(\sum_{i=1, i \neq y}^k u_i + \bar u_{-y} \right) \right)_+\\ 
&= \left(1 - (\bar u_{-y} + 1) + \frac 1 k \left((k-1) \bar u_{-y} + \bar u_{-y} \right) \right)_+\\
&= \left(-\bar u_{-y} + \bar u_{-y}\right)_+ \\
&= 0
\end{align*}

Now, for any outcome $i \neq y$, we know that $\bar u_{-i} > \bar u_{-i}$ because of $u_y$.
Since $u_y$ is only added in $L^{top-k}(\cdot)_i$ as part of the $\bar u_{-i}$ term, we can see that $L^{top-k}(u)_i \geq L^{top-k}(u')_i$ for all $i \neq y$.
\end{proof}

@@

First, observe that we can re-write
\begin{align*} 
L^{top-k}(u)_y&=\left(\frac{1}{k} \sum_{i=1}^k (u + \ones - e_y)_{[i]} - u_y \right)_+\\ 
&=\left(1 - u_y + \frac{1}{k} \sum_{i=1}^k (u - e_y)_{[i]} \right)_+\\
&=\left(1 - u_y + \frac{1}{k} \sum_{i=1}^k (u - \min(u_y, 1)\Ind{y \leq k})_{i} \right)_+
\end{align*}

Consider that the loss value is $1 + \bar u$ when $y > k$, since $u_y = 0$ in this case, so the loss is nonnegative even before applying the $(\cdot)_+$ operator.
If $y \leq k$ and $u_y \in [0,1]$, then the loss is nonnegative before applying $(\cdot)_+$ as $1-u_y \geq 0$, as is $\bar u$.
Now if $u_y \in (1, \bar u_{-y} + 1]$, then the term inside $(\cdot)_+$ of $L^{top-k}(u)_y$ is equal to $\frac{1}{k-1} + 1 +  \ubar_{-y} - u_y $
Since $u_y \leq 1 +  \ubar_{-y}$, we can see this term is nonnegative.
Therefore, we can drop the $(\cdot)_+$ operator on this report space.

We can now write the expected loss as follows:
\begin{align*}
\inprod{p}{L^{top-k}(u)} &= \sum_{y=1}^n p_y L^{top-k}(u)_y \\
&= \sum_{y=1}^b p_y L^{top-k}(u)_y + \sum_{y=b+1}^k p_y L^{top-k}(u)_y + \sum_{y=k+1}^n p_y L^{top-k}(u)_y\\
&= \sum_{y=1}^b p_y (1 - \frac 1 k - u_y + \ubar) + \sum_{y=b+1}^k p_y (1 - \frac{k+1} k u_y + \ubar) + \sum_{y=k+1}^n p_y(1 - u_y + \ubar)\\
&= 1 + \ubar - \inprod{p}{u} - \inprod{p \odot \Ind{y \leq b}}{\frac 1 k \ones} - \inprod{p \odot \Ind{b+1 \leq y \leq k}}{ \frac u k}\\
&= 1 + \ubar - \inprod{p}{u + \frac 1 k \min(u, \ones)}
\end{align*}
When considering the Bayes Risk, we use this final form of the expected loss.

\subsection{Studying the Bayes Risk}
First, consider the case where $u_y = 0$ and $u'_y = 1$, with $u_j = u'_j \;\forall i \neq j$.
As the expected loss is linear on $u_y$, and these are the boundary cases, we know that $u$ is optimal on one of these two reports in this case.

So we ask when is $\inprod{p}{L^{top-k}(u)} \leq \inprod{p}{L^{top-k}(u')}$?
\begin{align*}
- \inprod{p}{u + \frac 1 k \min(u, \ones)} &\leq \frac {1}{k} - \inprod{p}{u' + \frac 1 k \min(u', \ones)}\\
0 &\leq \frac{1}{k} - p_y(\frac{k+1}{k})\\
p_y &\leq \frac{1}{k+1}
\end{align*}

Conversely, if $p_y > \frac{1}{k+1}$, it is to better to report $u_y = 1$ than $u_y = 0$.
However, since we are not restricted to the unit hypercube, we might ask when it is optimal to bring $u_y = 1$ up to $u'_y = 1 + \epsilon$ given $\epsilon > 0$.
Here, we consider when $\inprod{p}{L^{top-k}(u)} \leq \inprod{p}{L^{top-k}(u')}$.
\begin{align*}
%1 + \ubar - \inprod{p}{u + \frac 1 k \min(u, \ones)} &\leq 1 + \ubar' - \inprod{p}{u' + \frac 1 k \min(u', \ones)}\\
\frac{1}{k} - p_y (\frac{k+1}{k}) &\leq \frac{1+\epsilon}{k} - p_y (\frac{k+1}{k}(1 + \epsilon))\\
p_y &\leq \frac{1}{k}
\end{align*}
In this case, $u_y=1$ is preferable to $u+y >1$.
If $p_y > \frac 1 k$, then we want to push $u_y$ as high as we can, but recall that we restrict to $u_y \leq \ubar_{-y} + 1$.
Because the upper bound on $u_y$ depends on $\ubar$, we can't determine $u$ iteratively.
In fact, we might have some $p_y \in [\frac 1 {k+1}, \frac 1 k]$ that we want to crank up past one, since this allows us to crank up other high-probability report values even farther.

\begin{theorem}
	For the loss $L^{top-k} : \reals^n \times \Y \to \reals_+$ given in Equation~\ref{eq:top-k-surrogate}, there is an optimal report in the following report set $\R$:
	\begin{align*}
	\R := \left\{ \frac{|M| + k -1}{k - |H|} \mathbf{1}_H + \mathbf{1}_M : H, M \subset [n], H\cap M = \emptyset, |H| + |M| \leq k \right\}
	\end{align*}
\end{theorem}
\begin{proof}[Proof Sketch.]
	The above work tells us that when $p_y > \frac 1 {k+1}$, it is optimal to set $u_y > 0$, and otherwise, it is optimal to set $u_y = 0$.
	Since $L^{top-k}$ is linear on $[0,1]$ and $[1, c]$ for some $c \geq 1$ and all dimensions $i \in [n]$, we know that the optimum is on the boundary of such reports, so when we bump $u_y$ up to some number greater than $0$, it should either be one $1$ or $c$ to be an optima.
	$H$ represents the elements of the report that are bumped up to $c$, and $M$ represents the elements of the report that are bumped up to $1$.
	Since we know that no distribution has more than $k$ outcomes with weight greater than $\frac 1 {k+1}$, we know that the size of $H \cup M$ must be less than or equal to $k$, and $H \cap M$ is empty by each element of $u$ being real-valued, so for $c > 1$, $u_i$ cannot simultaneously be equal to $c$ and $1$.
	(Note: if $c = 1$, then WLOG, we can say $H = \emptyset$.)

	To see the reason for $c := \frac{|M| + k -1}{k - |H|}$, consider the following: given the loss $L^{top-k}$, when we want to ``crank $u_i$'' above $1$, we want to do so as much as possible, until hitting some threshold discussed in Lemma~\ref{lem:top-threshold}, due to the $(\cdot)_+$ operator surrounding the loss.
	Exceeding the threshold on $u_i > c$ then can only increase expected loss.

	We can take $H$ to be the set of elements of $u$ that get ``cranked up,'' and $M$ to be the set of elements of $u$ that should be set to $1$.

	Then, for $i \in H$, cranking $u_i$ as high as possible, we want to have $c = u_i = \bar u _{-i} + 1$, which we can see that 
	\begin{align*}
	c &= \bar u_{-i} + 1 \\
	c &= \frac{(c (|H|-1)) + |M|}{k-1} + 1\\
	ck-c &= c|H| - c + |M| + k - 1\\
	ck - c|H| &= |M| + k - 1\\
	c &= \frac{|M|+k-1}{k-|H|}
	\end{align*}
\end{proof}

By the above, we now have a finite set of optimal reports for $L^{top-k}$.
This yields the finite loss $L|_\R$ embedded by $L^{top-k}$.

%\begin{conjecture}
%	Fix $p \in \simplex$.
%	Let $g_t = |\{p_i : p_i \geq \frac 1 k\}|$ and $g_s = | \{p_i \in (\frac{1}{k+1}, \frac{1}{k})\}|$.
%	The property \[\gamma(p) = 
%	\left[
%	\begin{cases}
%	0 & p_i \leq \frac {1}{k+1}\\
%	1 & p_i \in (\frac{1}{k+1}, \frac{1}{k})\\
%	\frac{g_s + k-1}{k-g_t} & o/w
%	\end{cases}
%	\right]_{i=1}^n
%	\] refines the property $\prop{L^{top-k}}$.
%\end{conjecture}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
